# Importing and transforming proteomics data

## Importing flat files

## Calculating fold change and enrichment

### Fold change and log-fold change

Fold changes are ratios, the ratio of say protein expression before and
after treatment, where a value larger than 1 for a protein implies that 
protein expression was greater after the treatment.

In life sciences, fold change is often reported as log-fold change. Why is that?
There are at least two reasons which can be shown by plotting.

One is that ratios are not symmetrical around 1, so it's difficult to observe
both changes in the forwards and backwards direcion i.e. proteins where expression
went up and proteins where expression went down due to treatment. When we 
transform ratios on a log scale, the scale becomes symmetric around 0 and thus
we can now observe the distribution of ratios in terms of positive, negative or
no change.

```{r fold-change-1}
set.seed(10)
x <- 2^(rnorm(100))
y <- 2^(rnorm(100))
ratios <- tibble(value = x/y, label = "ratios")
logratios <- tibble(value = log2(ratios$value), label = "logratios")

bind_rows(ratios,logratios) %>% 
  ggplot(aes(value)) +
  geom_histogram(binwidth = 2, colour = "grey50", fill = "white") +
  ggplot2::facet_wrap(~ label) +
  theme_minimal()
```

A second reason is that transforming values onto a log scale changes where
the numbers actually occur when plotted on that scale. If we consider the log
scale to represent magnitudes, then we can more easily see changes of small and
large magnitudes when we plot the data.

For example, a fold change of 32 times can be either a ratio 1/32 or 32/1. 

1/32 is much closer to 1 than 32/1, but transformed to a log scale we see that in 
terms of magnitude of difference it is the same as 32/1.

```{r fold-change-2}
x2 <- 2^seq(1,5)
y_vals <- c(rev(1/x2),1,x2)
names <- c(paste0("1/",rev(x2)),1,x2)
x_vals <- seq(along=y_vals)

dat <- tibble(x_vals,y_vals,names)

p1 <- ggplot(dat,aes(x_vals,y_vals, label = names)) +
  geom_text() +
  geom_hline(yintercept = 1) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(breaks = NULL)


p2 <- ggplot(dat,aes(x_vals,y_vals, label = names)) +
  geom_text() +
  geom_hline(yintercept = 1) +
  scale_y_continuous(trans = "log2") +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(breaks = NULL)

plot_grid(p1,p2)

```

# Fold change and t-test

1. Load the data, we need multiple replicates for each condition. To be tidy
each protein is a set of observations (rows) of the variables, which are the
values recorded for each replicate (columns).

2. Deal with missing values, either impute or exclude missing values

3. Use `t.test` to perform Welch Two Sample t-test. This output will include
the p-values we need for each protein.

4. Perform log transformation of the observations for each protein.

5. Calculate the mean observation for each protein under each condition.

6. The log fold change is then the difference between condition 1 and condition 2.
Plot a histogram to look at the distribution.

7. Create a combined table of log fold change and p-values for all the proteins
for plotting a volcano plot.
