# Importing and transforming proteomics data

## Importing flat files

## Calculating fold change and enrichment

### Fold change and log-fold change

Fold changes are ratios, the ratio of say protein expression before and
after treatment, where a value larger than 1 for a protein implies that 
protein expression was greater after the treatment.

In life sciences, fold change is often reported as log-fold change. Why is that?
There are at least two reasons which can be shown by plotting.

One is that ratios are not symmetrical around 1, so it's difficult to observe
both changes in the forwards and backwards direcion i.e. proteins where expression
went up and proteins where expression went down due to treatment. When we 
transform ratios on a log scale, the scale becomes symmetric around 0 and thus
we can now observe the distribution of ratios in terms of positive, negative or
no change.

```{r fold-change-1}
set.seed(10)
x <- 2^(rnorm(100))
y <- 2^(rnorm(100))
ratios <- tibble(value = x/y, label = "ratios")
logratios <- tibble(value = log2(ratios$value), label = "logratios")

bind_rows(ratios,logratios) %>% 
  ggplot(aes(value)) +
  geom_histogram(binwidth = 2, colour = "grey50", fill = "white") +
  ggplot2::facet_wrap(~ label) +
  theme_minimal()
```

A second reason is that transforming values onto a log scale changes where
the numbers actually occur when plotted on that scale. If we consider the log
scale to represent magnitudes, then we can more easily see changes of small and
large magnitudes when we plot the data.

For example, a fold change of 32 times can be either a ratio 1/32 or 32/1. 

1/32 is much closer to 1 than 32/1, but transformed to a log scale we see that in 
terms of magnitude of difference it is the same as 32/1.

```{r fold-change-2}
x2 <- 2^seq(1,5)
y_vals <- c(rev(1/x2),1,x2)
names <- c(paste0("1/",rev(x2)),1,x2)
x_vals <- seq(along=y_vals)

dat <- tibble(x_vals,y_vals,names)

p1 <- ggplot(dat,aes(x_vals,y_vals, label = names)) +
  geom_text() +
  geom_hline(yintercept = 1) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(breaks = NULL)


p2 <- ggplot(dat,aes(x_vals,y_vals, label = names)) +
  geom_text() +
  geom_hline(yintercept = 1) +
  scale_y_continuous(trans = "log2") +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(breaks = NULL)

plot_grid(p1,p2)

```

# Fold change and t-test

1. Load the data, we need multiple replicates for each condition. To be tidy
each protein is a set of observations (rows) of the variables, which are the
values recorded for each replicate (columns).

```{r load-data, echo=FALSE, eval=FALSE}
# Emily's sample data set, missing values are NaN and clean up names
dat_em <- read_csv("data/MergedFinalProt_1colPerFile_6files_filtered.csv", 
                   na = "NaN") %>% clean_names()

# Have a look at the dataset
glimpse(dat_em)
```

2. Tidy up and deal with missing values, either impute or exclude missing values

```{r tidy-data}
# Tidy data up
dat_tidy <- dat_em %>%
  # Remove missing values
  drop_na() %>% 
  # Select and rename columns
  select(protein_accession = protein_key_protein_accession,
         protein_description,
         wt1 = protein_ngram_on_column_emilly_bowler_wt_rep1_001_merged_spectrum_ia_final_protein,
         wt2 = protein_ngram_on_column_emilly_bowler_wt_rep2_001_merged_spectrum_ia_final_protein,
         wt3 = protein_ngram_on_column_emilly_bowler_wt_rep3_001_merged_spectrum_ia_final_protein,
         kog1 = protein_ngram_on_column_emilly_bowler_kogsk_rep1_001_merged_spectrum_ia_final_protein,
         kog2 = protein_ngram_on_column_emilly_bowler_kogsk_rep2_merged_ia_final_protein,
         kog3 = protein_ngram_on_column_emilly_bowler_kogsk_rep3_001_merged_spectrum_ia_final_protein)

# Normalise to maximum column value
dat_norm_max <- dat_tidy %>% 
  mutate(wt1 = wt1/max(wt1),
         wt2 = wt2/max(wt3),
         wt3 = wt3/max(wt3),
         kog1 = kog1 / max(kog1),
         kog2 = kog2 / max(kog2),
         kog3 = kog3 / max(kog3)
         )

# Normalise to median column value
dat_norm_med <- dat_tidy %>% 
  mutate(wt1 = wt1/median(wt1),
         wt2 = wt2/median(wt3),
         wt3 = wt3/median(wt3),
         kog1 = kog1 / median(kog1),
         kog2 = kog2 / median(kog2),
         kog3 = kog3 / median(kog3)
         )

# Have a look at the median normalised data
glimpse(dat_norm_med)

# Save the median normalised data
write_excel_csv(dat_norm_med,"data/example_median_nomralised_proteomics_data.csv")

```


```{r impute}
# Impute NA with lowest value, need to change this!
control <- dat_select %>% filter(!is.na(control_1) | 
                                   !is.na(control_2) | 
                                   !is.na(control_3)) %>% select(3,4,5) 

treatment <- dat_select %>% filter(!is.na(ras_1) | 
                                   !is.na(ras_2) | 
                                   !is.na(ras_3)) %>% select(6,7,8) 

# Impute minimum
row_min <- function(x) apply(x,1,min,na.rm = TRUE)
impute_min <- function(x) replace(x, is.na(x), row_min(x))

#control_na <- control %>% filter(!complete.cases(.))
impute_min(control)

 
control %>% replace_na(apply(.,1,min))
```

3. Use `t.test` to perform Welch Two Sample t-test on untransformed data. 
This outputs the p-values we need for each protein.


```{r t-test}
# T-test function for multiple experiments
expriments_ttest <- function(dt,grp1,grp2){
  # Subset control group and convert to numeric
  x <- dt[grp1] %>% unlist %>% as.numeric()
  # Subset treatment group and convert to numeric
  y <- dt[grp2] %>% unlist %>% as.numeric()
  # Perform t-test
  result <- t.test(x, y)
  # Return p-values
  return(result$p.value)
}  

# Apply t-test function to data
# array = dat, 1 = rows, FUN = expriments_ttest, and arguements
# For median normalised data
p_vals_med <- apply(dat_norm_med,1,expriments_ttest, grp1 = c(3:5), grp2 = c(6:8))
# For maximum normalised data
p_vals_max <- apply(dat_norm_max,1,expriments_ttest, grp1 = c(3:5), grp2 = c(6:8))

# Plot histograms
hist(p_vals_med)
hist(p_vals_max)
```

4. Perform log transformation of the observations for each protein.

```{r log-data}
# Select columns and log data
dat_med_log <- dat_norm_med %>% 
  select(-c(protein_accession,protein_description)) %>% log2()
dat_max_log <- dat_norm_max %>% 
  select(-c(protein_accession,protein_description)) %>% log2()
```

5. Calculate the mean observation for each protein under each condition.

```{r mean-log}
con_med <- apply(dat_med_log[,1:3],1,mean)
trt_med <- apply(dat_med_log[,4:6],1,mean)

con_max <- apply(dat_max_log[,1:3],1,mean)
trt_max <- apply(dat_max_log[,4:6],1,mean)
```

6. The log fold change is then the difference between condition 1 and condition 2.
Plot a histogram to look at the distribution.

```{r fold-change}
# Calculate fold change
dat_med_fc <- con_med - trt_med
dat_max_fc <- con_max - trt_max

# Plot histograms
hist(dat_med_fc)
hist(dat_max_fc)
```

7. Create a combined table of log fold change and p-values for all the proteins
for plotting a volcano plot.

```{r vol-pot}
dat_med <- data.frame(prots= dat_norm_med$protein_accession,
                      logfc = dat_med_fc, 
                      pval = -1*log10(p_vals_med))

dat_max <- data.frame(prots= dat_norm_max$protein_accession,
                      logfc = dat_max_fc, 
                      pval = -1*log10(p_vals_max))

dat_max %>% ggplot(aes(logfc,pval)) + geom_point()

dat_med %>% ggplot(aes(logfc,pval)) + geom_point()

```
