# Going further {#going-further}

Here are a few links and suggestions about what else you might like to do with
R.

## Exporting figures {#export-figs}

Exporting figures can be done using the following structure:

```{r export-plots, eval=FALSE}
# Open up a blank plot file, pdf,jpeg etc.
<plot_function("file",...)>
# Write the plot to the file
<plot_object>
# Close the file
dev.off()
```

For example to export the volcano plot from Figure \@ref(fig:nice-vplot) to a pdf, we do:

```{r export-pdf, eval=FALSE}
# Open up a blank plot file, pdf,jpeg etc.
pdf("volcano_plot.pdf")

# Write the plot to the file
dat_tf %>%
  # Add a threhold for significant observations
  mutate(threshold = if_else(log_fc >= 2 & log_pval >= 1.3 |
                               log_fc <= -2 & log_pval >= 1.3,"A", "B")) %>%
  # Plot with points coloured according to the threshold
  ggplot(aes(log_fc,log_pval, colour = threshold)) +
  geom_point(alpha = 0.5) + # Alpha sets the transparency of the points
  # Add dotted lines to indicate the threshold, semi-transparent
  geom_hline(yintercept = 1.3, linetype = 2, alpha = 0.5) + 
  geom_vline(xintercept = 2, linetype = 2, alpha = 0.5) +
  geom_vline(xintercept = -2, linetype = 2, alpha = 0.5) +
  # Set the colour of the points
  scale_colour_manual(values = c("A"= "red", "B"= "black")) +
  xlab("log2 fold change") + ylab("-log10 p-value") + # Relabel the axes
  theme_minimal() + # Set the theme
  theme(legend.position="none") # Hide the legend

# Close the file
dev.off()
```

If I had saved the plot to an object called `vplot` I would call that object
instead of making the plot using `dat_tf` as shown here.

[Here](https://www.stat.berkeley.edu/classes/s133/saving.html) is a general
guide to the various formats you can export to.

Alternatively, if you are working in `ggplot` you can use the `ggsave` function
as described in [R4DS 28.7](http://r4ds.had.co.nz/graphics-for-communication.html#saving-your-plots).

```{r ggsave, eval=FALSE}
# Plot
dat_tf %>%
  # Add a threhold for significant observations
  mutate(threshold = if_else(log_fc >= 2 & log_pval >= 1.3 |
                               log_fc <= -2 & log_pval >= 1.3,"A", "B")) %>%
  # Plot with points coloured according to the threshold
  ggplot(aes(log_fc,log_pval, colour = threshold)) +
  geom_point(alpha = 0.5) + # Alpha sets the transparency of the points
  # Add dotted lines to indicate the threshold, semi-transparent
  geom_hline(yintercept = 1.3, linetype = 2, alpha = 0.5) + 
  geom_vline(xintercept = 2, linetype = 2, alpha = 0.5) +
  geom_vline(xintercept = -2, linetype = 2, alpha = 0.5) +
  # Set the colour of the points
  scale_colour_manual(values = c("A"= "red", "B"= "black")) +
  xlab("log2 fold change") + ylab("-log10 p-value") + # Relabel the axes
  theme_minimal() + # Set the theme
  theme(legend.position="none") # Hide the legend

# Use ggsave to save the plot as pdf
ggsave("volcano_plot.pdf", width = 20, height = 20, units = "cm")
```


## Exporting data

There is a [full manual for the import and export of data](https://cran.r-project.org/doc/manuals/r-release/R-data.html) in R.
However here are few pointers:

### Writing to a file

One of the most portables way to share data is by writing to a csv file. These
files can be opened in many programs. The tidyverse package contains two
functions for csv files, `write_csv` and for Excel `write_excel_csv`. The
latter form adds a bit of metadata that tells Excel about the file encoding.
See [R4DS writing to a file](http://r4ds.had.co.nz/data-import.html#writing-to-a-file).

For example to write a csv file of `dat_tf` to a file called
`04072018_transformed_data.csv` to our working directory for sharing
with a colleague using excel, the code is of the form 
`<function>(<r-object>,"filename")` like so:

```{r write-file, eval=FALSE}
write_excel_csv(dat_tf,"04072018_transformed_data.csv")
```

Note that the file name is a string and is in quotes.

### For R

If you are exporting data to use yourself in R, the custom `.rds` format is a 
good choice and preserves the R structure.

In the tidyverse `write_rds` follows the same structure as `write_csv`.

You can read back in using `read_rds`.

## Joining the R community

It's worth joining the [RStudio Community](https://community.rstudio.com/) and
following community members on Twitter such as [Jenny Bryan](https://twitter.com/JennyBryan), [Hadley Wickham](https://twitter.com/hadleywickham), [Yihui Xie](https://twitter.com/xieyihui), [Mara Averick](https://twitter.com/dataandme),
[David Robinson](https://twitter.com/drob)
 and [Julia Silge](https://twitter.com/juliasilge).
 
 I would not reccomend DataCamp due to their [misconduct](https://juliasilge.com/blog/datacamp-misconduct/). 
 
 And [swirl](https://swirlstats.com/) is free.

## Communication: creating reports, presentations and websites

[R Markdown](https://rmarkdown.rstudio.com/lesson-1.html) [@R-rmarkdown] enables us to do
[literate programming](https://en.wikipedia.org/wiki/Literate_programming), saving time as we can create analysis, reports, dashboards or web apps at the same time
as writing code. R Markdown can use multiple programming languages. 
See also [R4DS R Markdown](http://r4ds.had.co.nz/r-markdown.html) and
[R4DS R Markdown formats](http://r4ds.had.co.nz/r-markdown-formats.html) and
the [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/).

As linked in Section \@ref(viz) Claus Wilke has written a very nice guide to 
visualising data using R cvisualising data using R called
[Fundamentals of Data Visualization](http://serialmentor.com/dataviz/index.html) 
which is very helpful when it comes to thinking about how to best create figures
for a report, poster or presentation.

You can use [blogdown](https://bookdown.org/yihui/blogdown/) to build websites.
I created this guide to [buidling an academic website with blogdown](http://ab604.github.io/docs/website_bookdown/).

### Using bookdown to write a thesis dissertaion 

I used the bookdown package to create these materials [@R-bookdown] and you can
use it to write a thesis dissertaion, as detailed very nicely in this blog by
[Edd Berry](https://eddjberry.netlify.com/post/writing-your-thesis-with-bookdown/).

## Machine Learning

If you are interested in machine learning, then [TensorFlow](https://tensorflow.rstudio.com/) is a good place to start, for 
example Leon Eyrich Jessen's [Deep Learning for Cancer Immunotherapy](https://tensorflow.rstudio.com/blog/dl-for-cancer-immunotherapy.html) tutorial.

## Version control

Another thing you may wish to consider is version control, *"a system that records changes to a file or set of files over time so that you can recall specific versions later".*

To get started, have a look at these [slides by Alice Bartlett ](https://speakerdeck.com/alicebartlett/git-for-humans) and check out the [Rstudio version control guide ](https://support.rstudio.com/hc/en-us/articles/200532077-Version-Control-with-Git-and-SVN).