# Going further {#going-further}

Here are a few links and suggestions about what else you might like to do with
R.

## Exporting data and figures

Exporting figures is best done using the following structure:

```{r export-plots, eval=FALSE}
# Open up a blank plot file, pdf,jpeg etc.
<plot_function("file",...)>
# Write the plot to the file
<plot_object>
# Close the file
dev.off()
```

For example to export the volcano plot from Figure \@ref(fig:nice-vplot) to a pdf, we do:

```{r export-pdf, eval=FALSE}
# Open up a blank plot file, pdf,jpeg etc.
pdf("volcano_plot.pdf")

# Write the plot to the file
dat_tf %>%
  # Add a threhold for significant observations
  mutate(threshold = if_else(log_fc >= 2 & log_pval >= 1.3 |
                               log_fc <= -2 & log_pval >= 1.3,"A", "B")) %>%
  # Plot with points coloured according to the threshold
  ggplot(aes(log_fc,log_pval, colour = threshold)) +
  geom_point(alpha = 0.5) + # Alpha sets the transparency of the points
  # Add dotted lines to indicate the threshold, semi-transparent
  geom_hline(yintercept = 1.3, linetype = 2, alpha = 0.5) + 
  geom_vline(xintercept = 2, linetype = 2, alpha = 0.5) +
  geom_vline(xintercept = -2, linetype = 2, alpha = 0.5) +
  # Set the colour of the points
  scale_colour_manual(values = c("A"= "red", "B"= "black")) +
  xlab("log2 fold change") + ylab("-log10 p-value") + # Relabel the axes
  theme_minimal() + # Set the theme
  theme(legend.position="none") # Hide the legend

# Close the file
dev.off()
```

If I had saved the plot to an object called `vplot` I would call that object
instead of making the plot using `dat_tf` as shown here.

[Here](https://www.stat.berkeley.edu/classes/s133/saving.html) is a general
guide to the various formats you can export to.

## Joining the R community

It's worth joining the [RStudio Community](https://community.rstudio.com/) and
following community members on Twitter such as [Jenny Bryan](https://twitter.com/JennyBryan), [Hadley Wickham](https://twitter.com/hadleywickham), [Yihui Xie](https://twitter.com/xieyihui), [Mara Averick](https://twitter.com/dataandme),
[David Robinson](https://twitter.com/drob)
 and [Julia Silge](https://twitter.com/juliasilge).
 
 If you can afford [DataCamp](https://www.datacamp.com) then this is my preferred
 learning platform.
 
 And if you can't, then [swirl](https://swirlstats.com/) is free.

## Communication: creating reports, presentations and websites

[R Markdown](https://rmarkdown.rstudio.com/lesson-1.html) enables us to do
[literate programming](https://en.wikipedia.org/wiki/Literate_programming), saving time as we can create analysis, reports, dashboards or web apps at the same time
as writing code. R Markdown can use multiple programming languages. 

You can use [blogdown](https://bookdown.org/yihui/blogdown/) to build websites.
I created s guide to [buidling an academic website with blogdown](http://ab604.github.io/docs/website_bookdown/).

## Machine Learning

If you are interested in machine learning, then [TensorFlow](https://tensorflow.rstudio.com/) is a good place to start, for 
example Leon Eyrich Jessen's [Deep Learning for Cancer Immunotherapy](https://tensorflow.rstudio.com/blog/dl-for-cancer-immunotherapy.html) tutorial.