--- 
title: "Data Science Workshop"
subtitle: "British Society for Proteomic Research Meeting 2018"
author: "Alistair Bailey"
date: "`r format(Sys.Date(),format='%B %d %Y')`"
output: 
  bookdown::gitbook: default
  bookdown::epub_book: default
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
    keep_tex: yes
    citation_package: natbib
    latex_engine: xelatex
documentclass: book
fontsize: 12pt
linestretch: 1.5
toc-depth: 1
secnumdepth: 1
lof: FALSE
lot: FALSE
site: bookdown::bookdown_site
bibliography: [packages.bib, book.bib]
biblio-style: apalike
link-citations: yes
#geometry: "left=4cm, right=3cm, top=2.5cm, bottom=2.5cm"
description: "Lessons for the BSPR 2018 Data Science Workshop"
---

```{r include=FALSE}
library(tidyverse)
library(janitor)
library(cowplot)
#library(VennDiagram)
#library(gridExtra)
#library(cowplot)
# automatically create a bib database for R packages
knitr::write_bib(c(.packages(), 'bookdown', 'knitr', 'rmarkdown'), 'packages.bib')

options(knitr.graphics.auto_pdf = TRUE)
```

# Overview {-}

These lessons cover:

1. An introduction to R and RStudio
2. An introduction to the tidyverse
3. Importing and transforming proteomics data
4. Visualisation of proteomics analysis

The analysis is of an example data set of observations for 7702 proteins
from cells in three control experiments and three treatment experiments.
The observations are signal intensity measurements from the mass spectrometer.
These intensities relate to the amount of protein in each experiment and under 
each condition. The analysis transforms the data to examine the effect of
treatment on the cellular proteome and visualise the output using a volcano
plot and a heatmap. 
<a href="https://raw.githubusercontent.com/ab604/ab604.github.io/master/docs/070718-proteomics-example-data.csv" download>Click here to download the csv file.</a>

## Requirements {-}

Up to date version of R [@R-base] and Rstudio [@rstudioteam2018]

The following R packages:

```{r packages, eval=FALSE}
install.packages(c("tidyverse","gplots","pheatmap"))
```



